---
title: "Getting started with msArrow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with msArrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>")
```

```{r setup}
library(msArrow)
library(dplyr)
```

## The idea

Mass spectrometry data is large enough that holding several intermediate
versions of it in the R session is the thing that actually runs you out of
memory. msArrow's answer is that **a variable holds a path, not the data**.

`write_data()` writes a table and returns the file it wrote. Every step
downstream re-reads that file.

```{r}
dir <- tempfile("msArrow_intro")
dir.create(dir)

data_raw <- tibble::tibble(
  Protein.Group = rep(c("P12345", "Q9Y6K9", "O00305"), each = 4),
  Condition     = rep(c("ctrl", "ctrl", "t8", "t8"), 3),
  Intensity     = c(10.1, 11.4, 15.2, 14.8,
                     5.5,  6.1,  5.9,  6.4,
                    22.0, 21.3, 30.7, 29.9)) %>%
  write_data(file = "data_raw", dir = dir)

data_raw
```

The variable is a path. To use it, read it:

```{r}
get_data(data_raw) %>%
  count(Protein.Group)
```

`get_data()` returns a materialised tibble, so it never needs a trailing
`collect()`.

## Reading lazily

When the file is large and you only need part of it, `open_data()` gives an
Arrow connection instead. The filter runs before anything is pulled into R.

```{r}
open_data(data_raw) %>%
  filter(Condition == "t8") %>%
  summarise(mean_int = mean(Intensity), .by = Protein.Group) %>%
  collect()
```

## Choosing the file type

Left alone, `write_data()` decides for itself: parquet when Arrow can
represent the object, a `.parquetlist` folder for a list, `.Rds` otherwise.

The `type` argument overrides that. It takes one or several of `parquet`,
`rds`, `tsv`, `csv` and `txt`.

```{r}
data_summary <- get_data(data_raw) %>%
  summarise(mean_int = mean(Intensity),
            n        = n(),
            .by      = c(Protein.Group, Condition)) %>%
  write_data(file = "data_summary", dir = dir, type = c("parquet", "tsv"))

list.files(dir, pattern = "data_summary")
```

Both files were written. **The returned path is the first type only**:

```{r}
basename(data_summary)
```

That is the point of the ordering. A pipeline stays on the fast parquet file,
while a `.tsv` that a collaborator can open lands beside it without a second
line of code. For a small table that only ever needs to be human-readable,
`type = "tsv"` on its own does the job:

```{r}
get_data(data_raw) %>%
  distinct(Protein.Group) %>%
  write_data(file = "protein_list", dir = dir, type = "tsv") %>%
  get_data()
```

`.txt` is written tab-separated, `.csv` comma-separated. Delimited files are
written through Arrow, and values are quoted where needed so a field
containing a tab or comma still round-trips:

```{r}
tibble::tibble(Genes = c("A,B", "C\tD"), n = 1:2) %>%
  write_data(file = "tricky", dir = dir, type = "tsv") %>%
  get_data()
```

Delimited formats store flat tables only. Asking for a `.tsv` of something
with a list column is an error rather than a silently mangled file.

## Skipping work that is already done

`write_data()` recalculates by default. Pass `redo = FALSE` to return the
existing path instead, when a step is genuinely expensive and you are sure the
inputs have not changed:

```{r}
write_data(get_data(data_raw), file = "data_raw", dir = dir, redo = FALSE)
```

With several types, the computation is skipped only when *every* requested
file already exists — otherwise the missing one would never be written.

## Lists become folders

A list is written as a folder of files rather than one large `.Rds`, so a
single element can be read without loading the rest.

```{r}
data_split <- get_data(data_raw) %>%
  split(.$Condition) %>%
  write_data(file = "by_condition", dir = dir)

list.files(data_split)

get_data(data_split)$t8
```

## Where the files went

Without a `file` name, `write_data()` writes into `tempdir()`. These helpers
show what has accumulated and clear it:

```{r}
length(tempdir_list())
```

```{r, include = FALSE}
unlink(dir, recursive = TRUE)
```

## Next steps

For annotated datasets — variables, observations and long
`observations x variables` frames kept together on disk, and DIA-NN reports
imported straight into that layout — see
[msTools](https://nicohuttmann.github.io/msTools/), which builds that store on
top of the functions shown here.
